Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2015-05-29 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Bench Scale Chapter complete, Wall Tests started

\documentclass[twoside]{uocthesis}

@@

Figure~\ref{demarcation} is a photograph of the edge of a fire pattern generated by the natural gas burner.  It will serve as visual definition of the line of demarcation to be used in this study to determine the limits of a fire pattern.  Referring to DeHaan's classifications, there are no surface deposits such as light soot that could be brushed or blown off the gypsum wallboard.  There is a thin band of surface thermal effects in the form of discoloration of the paint, this could be considered an outer bound of the pattern. To the left of the discoloration is clearly an larger area where charring (blackening of the paint/paper), penetration (lifting of the painted/paper residue), and consumption (areas where the paint/paper reside is gone altogether).  The last two classifications or criteria will be used in this study to define the edge of the fire pattern.  Any area where the paper has been burned away or penetrated such that it has lifted away from the gypsum will be considered part of the fire pattern and define the lines of demarcation for this study.

@@

 The initial compartment experimental series were conducted to examine the impact of the method of construction of the walls on the development of the fire patterns. These experiments would determine how the walls for the remaining compartment fire pattern experiments would be constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3)	Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting,a kraft paper facing, and an ``R value'' of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

The burner was centered on the middle panel of the rear wall and was in contact with the wall. The wall behind the center panel of the rear wall was constructed either as open, closed, or insulated to meet the needs of the experiment. The burner was positioned against the face of the painted gypsum board. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~s burn time which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.

@@

@@

Figure~\ref{NG_Open_Wall} shows a comparison of the ten fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sized however they are scaled so that the images are comparable.
The grids were used to assist with the determination of the fire pattern area and the sizing of the images. In order to measure the dimensions of the fire pattern, such as height and width, a determination had to be made as to where the fire pattern stops.  The boundary between the undamaged area and the area that has been damaged by the fire is known as the line of demarcation~\cite{NFPA_921}.

@@

@@

>>>>

@manual{ID,
	ALTauthor = {author},
	ALTeditor = {editor},
	title = {title},
	date = {date},
	OPTsubtitle = {subtitle},
	OPTtitleaddon = {titleaddon},
	OPTlanguage = {language},
	OPTedition = {edition},@unpublished{ID,
		author = {author},
		title = {title},
		date = {date},
		OPTsubtitle = {subtitle},
		OPTtitleaddon = {titleaddon},
		OPTlanguage = {language},
		OPThowpublished = {howpublished},
		OPTnote = {note},
		OPTlocation = {location},
		OPTisbn = {isbn},
		OPTdate = {date},
		OPTmonth = {month},
		OPTyear = {year},
		OPTaddendum = {addendum},
		OPTpubstate = {pubstate},
		OPTurl = {url},
		OPTurldate = {urldate},
	}

	OPTtype = {type},
	OPTseries = {series},
	OPTnumber = {number},
	OPTversion = {version},
	OPTnote = {note},
	OPTorganization = {organization},
	OPTpublisher = {publisher},
	OPTlocation = {location},
	OPTisbn = {isbn},
	OPTchapter = {chapter},
	OPTpages = {pages},
	OPTpagetotal = {pagetotal},
	OPTaddendum = {addendum},
	OPTpubstate = {pubstate},
	OPTdoi = {doi},
	OPTeprint = {eprint},
	OPTeprintclass = {eprintclass},
	OPTeprinttype = {eprinttype},
	OPTurl = {url},
	OPTurldate = {urldate},
}
\documentclass[twoside]{uocthesis}

@@

Figure~\ref{demarcation} is a photograph of the edge of a fire pattern generated by the natural gas burner.  It will serve as visual definition of the line of demarcation to be used in this study to determine the limits of a fire pattern.  Referring to DeHaan's classifications, there are no surface deposits such as light soot that could be brushed or blown off the gypsum wallboard.  There is a thin band of surface thermal effects in the form of discoloration of the paint, this could be considered an outer bound of the pattern. To the left of the discoloration is clearly an larger area where charring (blackening of the paint/paper), penetration (lifting of the painted/paper residue), and consumption (areas where the paint/paper reside is gone altogether).  The last three classifications or criteria will be used in this study to define the edge of the fire pattern.  Any area where the painted paper surface has been charred, penetrated such that it has lifted away from the gypsum core, or burned away will be considered part of the fire pattern and define the lines of demarcation for this study.

@@

 The initial set of compartment experiments were conducted to examine if the type of wall construction would impact the development of the fire patterns. These experiments would determine how the walls for the remaining compartment fire pattern experiments would be constructed.  Three cases were examined: 1) Open back, 12.7~mm thick, regular core, painted gypsum board on the front of a wood frame only, 2) Closed back, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame representative of an interior wall, and 3) Insulated, 12.7~mm thick, regular core, painted gypsum board on the front and back of a wood frame, with insulation in the voids of the wood frame representative of an exterior wall. The insulation was 90~mm thick fiber glass batting,a kraft paper facing, and an ``R value'' of 13.  From the instrumented gypsum wallboard experiments, the peak temperatures on the backside of the wallboard occurred when the burners were positioned against the wall.  The peak temperatures were similar for all of the fuels, approximately 80~$^\circ$C, although the duration of the higher temperatures on the rear of the wallboard was longer for the natural gas and the gasoline fueled burners. Therefore the wall construction experiments were only conducted with the natural gas and gasoline fueled burners as the impact of insulating the back of the walls exposed to the fuels with the highest energy would produce the most significant differences, if any, to the fire patterns.

@@

The burner was centered on the middle panel of the rear wall and was in contact with the wall. The wall behind the center panel of the rear wall was constructed either as open, closed, or insulated to meet the needs of the experiment. The burner was positioned against the face of the painted gypsum wallboard. As in the fire source characterization experiments the top of the fuel in each case was positioned 7.5~cm above the floor of the compartment. The source fires remained the same in terms of fuels, amounts of fuels used, burner size and burn duration.  The natural gas had a 300~s burn time which was approximately the upper bound for the total burn time for the 500~mL of gasoline.  By using the same burners, these compartment fire pattern experiments can be correlated with the instrumented wall experimental results.

@@

For this analysis, the outline of the fire pattern, or the line of demarcation~\cite{NFPA_921}, was defined as the points where charring, penetration and consumption of the paper covering the gypsum wallboard stopped. The grids were used to assist with the determination of the fire pattern area and the sizing of the images. This defining edge was used to measure the dimensions of the fire pattern, such as height and width. Given the irregular shape of the patterns, the area of each pattern could not be calculated based solely on the maximum height and maximum width measurements. Therefore the areas of the grid cells within the fire pattern were added up to determine the area of the fire patterns.  The analysis was carried out with a resolution of a quarter grid cell or approximately 0.003~m$^2$.

There were four measures that were taken to characterize each pattern: maximum height, maximum width, height at maximum width, and the area.  Each dimensional value is presented with the 95~\% confidence level (2~$\sigma$) based on Type A statistical analysis of the fire pattern measurements.

There are also a number of qualitative terms from NFPA~921 that were used to describe features of the patterns.  For example, "clean burn" to describe a lack of soot within a fire pattern area or "inverted cone" to describe the shape of the pattern~\cite{NFPA_921}.

@@

Figure~\ref{NG_Open_Wall} shows a comparison of the ten fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sized in an effect to make the images comparable.
The average maximum height of the fire pattern generated by the natural gas burner on the open walls was 0.74~m~$\pm~0.12~\%$.  The average maximum width of the fire pattern from the same experiments was 0.24~m~$\pm~0.06~\%$.  The position from the top edge of the burner to the vertical position of the maximum width was 0.41~m~$\pm~0.07~\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.05~\%$.

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each patterns is similar to the width of the burner. Moving up from the burner, about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and combusts, the plume and the pattern expands.  As the flames move up the wall the width tends to increase until about half of the peak height of the pattern, then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame is reduced.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticable bias to one side or the other can be seen in four of the ten experiments.

The fire pattern photographs from the ten closed wall construction experiments are shown in Figure~\ref{NG_Closed_Wall}. 
The patterns appear very similar in nature and size to the patterns that were generated on the open construction walls.  The average maximum height of the fire patterns generated by the natural gas burner on the closed construction wall was the same as the patterns generated on the open construction wall, 0.74~m~$\pm~0.06~\%$.  The 95~$\%$ confidence limit is about half of the value from the open wall experiments. The average maximum width of the fire patterns was 0.24~m~$\pm~0.02~\%$.  Again similar to the previous experiments. The vertical position of the maximum width was 0.43~m~$\pm~0.03~\%$. The area of the burn estimated by examining the damaged grid cells was 0.14~$m^2~\pm~0.01~\%$.

The basic shape of these patterns on the closed wall are very similar to the patterns on the open wall.  Six of the ten patterns have a bump out or an unsymmetrical curve within 200~mm of the top of the burner.  The cause of these bumps in the patterns could not be determined. Three of the patterns have a noticeable lean.

The last set of wall experiments with the natural gas burner was conducted with the insulated wall.  The patterns from the ten replicate experiments are shown in Figure~\ref{NG_Insulated_Wall}.  The average maximum height of the fire patterns generated by the natural gas burner on the insulated walls was 0.71~m~$\pm~0.10~\%$.  The average maximum width of the fire patterns on the insulated walls were 0.23~m~$\pm~0.04~\%$.  The vertical position of the maximum width was 0.42~m~$\pm~0.07~\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.02~\%$.

For all intents and purposes, the patterns generated on the three different types of wall construction were the same.  The comparison of the numerical values can be seen in Table~\ref{tab:Fire_Pattern_Dimensions_Wall_Construction}.  Each of the sets of 
patterns have a few patterns with the bump on one side of the pattern within 200~mm of the burner, as discussed in the section on the closed wall construction.

@@

Figure~\ref{Gas_Open_Wall} shows a comparison of the ten fire patterns generated by the natural gas burner on the wall with open back construction.  The images within the figure are slightly different sized in an effect to make the images comparable.
The average maximum height of the fire pattern generated by the natural gas burner on the open walls was 0.74~m~$\pm~0.12~\%$.  The average maximum width of the fire pattern from the same experiments was 0.24~m~$\pm~0.06~\%$.  The position from the top edge of the burner to the vertical position of the maximum width was 0.41~m~$\pm~0.07~\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.05~\%$.

These fire patterns, as well as all of the other patterns in this study, are fire plume generated patterns. The bottom width of each patterns is similar to the width of the burner. Moving up from the burner, about 25~mm, the pattern width is reduced as a result of air entrainment into the plume.  As the oxygen in the air mixes with the gaseous fuel and combusts, the plume and the pattern expands.  As the flames move up the wall the width tends to increase until about half of the peak height of the pattern, then the pattern starts to reduce in width as the fuel is consumed and the volume of the flame is reduced.  Based on observation alone, the visible flames extend well above the height of the resulting patterns.  Even though the experiments were conducted under quiescent conditions, a noticable bias to one side or the other can be seen in four of the ten experiments.

The fire pattern photographs from the ten closed wall construction experiments are shown in Figure~\ref{Gas_Closed_Wall}. 
The patterns appear very similar in nature and size to the patterns that were generated on the open construction walls.  The average maximum height of the fire patterns generated by the natural gas burner on the closed construction wall was the same as the patterns generated on the open construction wall, 0.74~m~$\pm~0.06~\%$.  The 95~$\%$ confidence limit is about half of the value from the open wall experiments. The average maximum width of the fire patterns was 0.24~m~$\pm~0.02~\%$.  Again similar to the previous experiments. The vertical position of the maximum width was 0.43~m~$\pm~0.03~\%$. The area of the burn estimated by examining the damaged grid cells was 0.14~$m^2~\pm~0.01~\%$.

The basic shape of these patterns on the closed wall are very similar to the patterns on the open wall.  Six of the ten patterns have a bump out or an unsymmetrical curve within 200~mm of the top of the burner.  The cause of these bumps in the patterns could not be determined. Three of the patterns have a noticeable lean.

The last set of wall experiments with the natural gas burner was conducted with the insulated wall.  The patterns from the ten replicate experiments are shown in Figure~\ref{Gas_Insulated_Wall}.  The average maximum height of the fire patterns generated by the natural gas burner on the insulated walls was 0.71~m~$\pm~0.10~\%$.  The average maximum width of the fire patterns on the insulated walls were 0.23~m~$\pm~0.04~\%$.  The vertical position of the maximum width was 0.42~m~$\pm~0.07~\%$. The area of the burn estimated by examining the damaged grid cells was 0.15~$m^2~\pm~0.02~\%$.

For all intents and purposes, the patterns generated on the three different types of wall construction were the same.  The comparison of the numerical values can be seen in Table~\ref{tab:Fire_Pattern_Dimensions_Wall_Construction}.  Each of the sets of 
patterns have a few patterns with the bump on one side of the pattern within 200~mm of the burner, as discussed in the section on the closed wall construction.

@@

Impact of Instrumentation on Walls?

Impact of Wall Construction
